### 2018-12-29 code.tex
finished results and analysis chapter

% Please add the following required packages to your document preamble:
% \usepackage{graphicx}
\begin{table}[h]
\centering
\resizebox{0.8\textwidth}{!}{%
\begin{tabular}{lr|r|r|r|r|r|r|r|}
\cline{3-9}
                             & \multicolumn{1}{l|}{}      & \multicolumn{7}{c|}{AROT [s]}                                                                                                                                                                      \\ \hline
\multicolumn{1}{|l|}{runway} & \multicolumn{1}{l|}{count} & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
\multicolumn{1}{|l|}{RWY 01} & 963 & 85 & 21 & 46 & 68 & 88 & 100 & 152 \\ \hline
\multicolumn{1}{|l|}{RWY 10} & 852 & 85 & 11 & 61 & 78 & 84 & 91 & 155 \\ \hline
\multicolumn{1}{|l|}{RWY 19} & 1511 & 67 & 10 & 46 & 61 & 66 & 71 & 155 \\ \hline
\multicolumn{1}{|l|}{RWY 28} & 401 & 68 & 14 & 49 & 60 & 66 & 72 & 155 \\ \hline
\end{tabular}%
}
\caption[AROTs for the air traffic mix by runway for the summer]{AROT statistics for the air traffic mix at KEF by runway for the summer of 2018. The count is the number of landings in peak hours.}
\label{my-label3}
\end{table}
\fxnote{Include a schematic of the airport, overview for runway and taxiway reference}

@@

\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{graphics/BIKF_schematic.png}
    \caption[BIKF schematic]{BIKF schematic with marked runways and fast-exit taxiways.}
    \label{fig:BIKF_schematic}
\end{figure}
% ----------------------------

>>>>

% Please add the following required packages to your document preamble:
% \usepackage{graphicx}
\begin{table}[h]
\centering
\resizebox{0.8\textwidth}{!}{%
\begin{tabular}{lr|r|r|r|r|r|r|r|}
\cline{3-9}
                             & \multicolumn{1}{l|}{}      & \multicolumn{7}{c|}{AROT [s]}                                                                                                                                                                      \\ \hline
\multicolumn{1}{|l|}{runway} & \multicolumn{1}{l|}{count} & \multicolumn{1}{l|}{mean} & \multicolumn{1}{l|}{std} & \multicolumn{1}{l|}{min} & \multicolumn{1}{l|}{25\%} & \multicolumn{1}{l|}{50\%} & \multicolumn{1}{l|}{75\%} & \multicolumn{1}{l|}{max} \\ \hline
\multicolumn{1}{|l|}{RWY 01} & 963 & 85 & 21 & 46 & 68 & 88 & 100 & 152 \\ \hline
\multicolumn{1}{|l|}{RWY 10} & 852 & 85 & 11 & 61 & 78 & 84 & 91 & 155 \\ \hline
\multicolumn{1}{|l|}{RWY 19} & 1511 & 67 & 10 & 46 & 61 & 66 & 71 & 155 \\ \hline
\multicolumn{1}{|l|}{RWY 28} & 401 & 68 & 14 & 49 & 60 & 66 & 72 & 155 \\ \hline
\end{tabular}%
}
\caption[AROTs for the air traffic mix by runway for the summer]{AROT statistics for the air traffic mix at KEF by runway for the summer of 2018. The count is the number of landings in peak hours.}
\label{my-label3}
\end{table}

@@

\chapter{Inter-arrival distance separation}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.5\textwidth]{graphics/fig_HM_to_DD_pairs_dist_separ.png}
    \caption[Inter-arrival distance separation of H-M pairs to D-D pairs into the RECAT scheme]{Inter-arrival distance separation after re-categorisation of H-H pairs ti D-D pairs into the RECAT scheme.}
    \label{fig:HM_to_DD_pairs_dist_separ}
\end{figure}

\chapter{BIKF Schemati}

\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{graphics/BIKF_schematic.png}
    \caption[BIKF schematic]{BIKF schematic with marked runways and fast-exit taxiways.}
    \label{fig:BIKF_schematic}
\end{figure}

### 2018-12-29 introduction.tex
finished results and analysis chapter

\subsection{Current Understandings of Wake Vortex Behaviour}
The wing of an aircraft can be regarded as a beam of finite length with aerofoil as cross section~\cite{hansen2015aerodynamics}\fxnote{Find a more appropriate source for citation or drop it if unnecessary. Read about induced vertical flow in a book on aerodynamics}.
The lift effect is created by the differential pressure between the lower and the upper side of the wing. At the wing tips the high pressure flow from the lower side leaks around the tip and to the upper side of the wing. Thus the streamlines over the wing are pushed inwards, while the streamlines under the wing are pushed outwards. A finite lifting wing leaves behind a continuous sheet of stream-wise vorticity ~\cite{hansen2015aerodynamics}\fxnote{Find a more appropriate source for citation }.  This manifests itself in multiple trailing swirls of the airflow that mix together and roll up into two main counter-rotating vortices aft of the wing as illustrated in Figure~\ref{fig:vortex_develop}~\cite{magazine_aibus_safety, Breitsamter2011Feb, gerz_commercial_2002}. Between the vortices the air moves downwards while outside the induced flow is upwards. Those vortices are also known as wake turbulence. 
The kinetic energy contained in the vortices is dependent on the weight and aerodynamics of the generating aircraft. The cross flow velocities in the core region of the trailing vortices can reach $360$ km/h and the vortices can stay effective up to hundred wing spans. This can result in wake vortices lasting for several minutes and up to $30$ km behind larger aircraft~\cite{Breitsamter2011Feb, gerz_commercial_2002}.

@@

Near ground on approach the wake vortices tend to descend slowly to an altitude of about one half of the initial separation.
Upon reaching this point the descent will stop and then ascend slowly. This "re-bounce" effect is caused by the presence of the ground and carries with it the formation of a second pair of induced vortices outside and bellow the main vortex. 
In the presence of stable cross-wind conditions the decay of the downwind vortex is not identical to the upwind vortex and the descent rate may vary significantly \cite{Hallock2018Apr}. 
The trajectories of the vortex pair may be modified and the upwind vortex could stall over the runway while the downwind vortex ascends and decays faster (Figure~\ref{fig:vortex_ground_effect}).
Measurements have shown that the lateral/sideways motion of a vortex in an airport environment could range between $229$~m and $518$~m and in some cases even $762$~m, which is the basis for the "2500-foot"  rule separation in parallel runways configuration~\cite{Hallock2018Apr, hallock2004summary, hallock2003wake}. The vertical descent rate of vortices for medium commercial aircraft in stable atmosphere  is shown to be around $1.5-2.5$~m/s for the first $30$~s after which the descent slows and eventually approaches zero at $152-274$ m below the flight path, and for heavier aircraft decaying vortices have been observed at $305$~m and below of the flight path~\cite{lissaman1973aircraft, Hallock2018Apr}. 
Aerodynamic properties of the aircraft govern the vortex roll-up process and ambient atmospheric conditions dominate the behaviour of the vortices forcing instability and eventual decay of the vortices~\cite{Hallock2018Apr}.
Several factors that drive the decay rate of the vortex pair have been formulated:
\begin{itemize}
    \item Atmospheric turbulence extracts energy from the vortex and reduces its strength leading to a faster wake decay ~\cite{Hallock2018Apr}.
    \item Viscosity of the atmosphere also draws out energy from the vortex but at a slower rate than the atmospheric turbulence~\cite{Hallock2018Apr}.\fxnote{This needs to be explained better (Ármann)}
    \item Buoyancy force acts on the vortex in a thermally stable stratified environment as a result of the lesser air density inside the vortex system and may causes stall or rebound to the flight level~\cite{Holzapfel2001Feb, gerz_commercial_2002}.
    \item Vortex instability in the form of long wave sinusoidal fluctuations of the vortex core (Crow instability) may occur due to light turbulence in the atmosphere and may cause the vortices to link and decay faster~\cite{Hallock2018Apr, crow2003stability}.
    \item Secondary vorticity structures, vertical rib-like counter rotating flow formations may appear between the vortex pair and eventually wrap around the main vortices, leading to turbulence build inside the system and rapid circulation decay~\cite{Holzapfel2001Feb, Holzapfel2003Jun}.
\end{itemize}

>>>>

\subsection{Current Understandings of Wake Vortex Behaviour}

The lift effect on wing is created by the differential pressure between the lower and the upper surface of the wing. At the wing tips the high pressure flow from the lower side leaks around the tip and to the upper side of the wing. Thus the streamlines over the wing are pushed inwards, while the streamlines under the wing are pushed outwards. When the two streams combine at the trailing edge of a lifting wing, the difference in span-wise velocity causes the air to roll up into a number of small stream-wise vortices distributed along the span of the wing that eventually mix together and combine into two main counter-rotating vortices aft of the wing as illustrated in Figure~\ref{fig:vortex_develop} ~\cite{houghton2012aerodynamics,magazine_aibus_safety, Breitsamter2011Feb, gerz_commercial_2002}. Between the vortices the induced flow is downwards (downwash) while outside the air moves upwards. Those vortices are also known as wake turbulence. 
The kinetic energy contained in the vortices is dependent on the weight and aerodynamics of the generating aircraft. The cross flow velocities in the core region of the trailing vortices can reach $360$ km/h and the vortices can stay effective up to hundred wing spans, which can result in wake vortices lasting for several minutes and up to $30$ km behind larger aircraft~\cite{Breitsamter2011Feb, gerz_commercial_2002}.

@@

Near ground on approach the wake vortices tend to descend slowly to an altitude of about one half of the initial separation.
Upon reaching this point the descent will stop and then ascend slowly. This "re-bounce" effect is caused by the presence of the ground and carries with it the formation of a second pair of induced vortices outside and bellow the main vortex. 
In the presence of stable cross-wind conditions the decay of the downwind vortex is not identical to the upwind vortex and the descent rate may vary significantly \cite{Hallock2018Apr}. 
The trajectories of the vortex pair may be modified and the upwind vortex could stall over the runway while the downwind vortex ascends and decays faster (Figure~\ref{fig:vortex_ground_effect}).
Measurements have shown that the lateral/sideways motion of a vortex in an airport environment could range between $229$~m and $518$~m and in some cases even $762$~m, which is the basis for the "2500-foot"  rule separation in parallel runways configuration~\cite{Hallock2018Apr, hallock2004summary, hallock2003wake}. The vertical descent rate of vortices for medium commercial aircraft in stable atmosphere  is shown to be around $1.5-2.5$~m/s for the first $30$~s after which the descent slows and eventually approaches zero at $152-274$ m below the flight path, and for heavier aircraft decaying vortices have been observed at $305$~m and below of the flight path~\cite{lissaman1973aircraft, Hallock2018Apr}. 
Aerodynamic properties of the aircraft govern the vortex roll-up process and ambient atmospheric conditions dominate the behaviour of the vortices forcing instability and eventual decay of the vortices~\cite{Hallock2018Apr}.
Several factors that drive the decay rate of the vortex pair have been formulated:
\begin{itemize}
    \item Atmospheric turbulence extracts energy from the vortex and reduces its strength leading to a faster wake decay ~\cite{Hallock2018Apr}.
    \item Viscosity of the atmosphere also draws out energy from the vortex but at a slower rate than the atmospheric turbulence. The so called dissipative action of viscosity  effectively removes energy from a disturbance, in this case a vortex, thereby causing it to decay~\cite{Hallock2018Apr, houghton2012aerodynamics}.
    \item Buoyancy force acts on the vortex in a thermally stable stratified environment as a result of the lesser air density inside the vortex system and may causes stall or rebound to the flight level~\cite{Holzapfel2001Feb, gerz_commercial_2002}.
    \item Vortex instability in the form of long wave sinusoidal fluctuations of the vortex core (Crow instability) may occur due to light turbulence in the atmosphere and may cause the vortices to link and decay faster~\cite{Hallock2018Apr, crow2003stability}.
    \item Secondary vorticity structures, vertical rib-like counter rotating flow formations may appear between the vortex pair and eventually wrap around the main vortices, leading to turbulence build inside the system and rapid circulation decay~\cite{Holzapfel2001Feb, Holzapfel2003Jun}.
\end{itemize}
